### 2022-02-14 manuscript/main.tex
Update manuscript

\textbf{Observations and analysis of an urban boundary layer during extreme heat events}
Gabriel Rios \textsuperscript{1*, 2*}, Prathap Ramamurthy \textsuperscript{1, 2}

@@

Extreme heat presents a significant risk to human health and infrastructure in cities. Several studies have been conducted in the past two decades to understand the interaction between the synoptic-scale extreme heat events and local-scale urban heat island effects. However, observations of boundary layer characteristics during these periods have been relatively rare. Our current understanding of boundary layer dynamics is incomplete, particularly in coastal urban environments where the local climatology is highly influenced by land-sea thermal gradients. Here we analyze the evolution and structure of the urban boundary layer during regular and extreme heat periods. Our primary goal is to understand how boundary layer dynamics during extreme heat events regulate near-surface transport of mass, momentum, and heat. The analysis will also focus on how the urban surface layer interacts with the mixed layer during these events. Our analysis focuses on the New York City metropolitan area and relies on observations from vertical profilers (Doppler lidar, microwave radiometer) and quantities derived by analytical methods. Additionally, satellite and reanalysis data are used to supplement observational data.

@@

\begin{enumerate}
  \item How do UBL structure and dynamics depart from the climatology during extreme heat events?
  \item How does the UBL structure impact the transport of scalars?
  \item What effect does the sea breeze have on a coastal urban area during extreme heat events?
\end{enumerate}

@@

\subsection{Temperature}
On average, extreme heat events increase the temperature at the surface, as expected (see Figure \ref{fig:extreme_heat_normal_0m_comparison_T}). This is consistent across all observed locations in New York City, with the extreme heat event temperature exceeding normal temperatures by approximately 1-$\sigma$ over the entire day. An increase in the difference is observed during daytime hours, with the difference peaking in magnitude around 13:00 LST at the hottest time of day. The surface temperature variability is significantly lower during heat events ($ \sigma = \SI{1.77}{\kelvin} $) than during normal temperatures ($ \sigma = \SI{4.57}{\kelvin} $).
\\ \\
Above the surface, extreme heat events similarly increase the temperature significantly over the lowest \SI{3000}{\meter} of the troposphere (see Figure \ref{fig:extreme-heat-normal-comparison-contours-theta}), with standardized anomalies of $\theta$ ranging from $\sigma = 0.88$ to $1.20$. The largest temperature anomalies shift from the surface layer in the mornings to span the entirety of the mixed layer. This is reflective of strong surface forcing resulting in convection through the surface layer, as indicated by the formation of a late morning superadiabatic layer at all locations (Figures \ref{fig:vertical_profiles-heat_wave-normal-theta-bron}, \ref{fig:vertical_profiles-heat_wave-normal-theta-quee}, \ref{fig:vertical_profiles-heat_wave-normal-theta-stat}). 
\\ \\
The vertical profiles of $\theta$ suggests a degree of spatial variability exists between locations. One instance of this spatial variability is vertical mixing; the Bronx site appears to have stronger vertical mixing based on Figure \ref{fig:vertical_profiles-heat_wave-normal-theta-bron}, as $\theta$ remains constant for a greater height than at the Queens and Staten Island locations. This phenomenon is more pronounced during extreme heat events, as a distinct mixed layer is apparent in the Bronx during early (12:00 LST) and late (18:00 LST) afternoon hours. While the mixed layer is also visible for the other locations, the strength of the mixed layer is emphasized by the negative $\frac{d\theta}{dz}$ values between 1000 and 1500 m.

\subsection{Moisture}
On average, extreme heat events were found to increase the moisture at the surface, as indicated by the diurnal profiles of specific humidity ($q$) (see Figure \ref{fig:extreme-heat-normal-0m-comparison-q}). This is also consistent across all observed locations in New York City, with the extreme heat event temperature exceeding normal temperatures by approximately 1-$\sigma$ over the entire day. Although a distinct diurnal profiles exists ($q$ decreases during daytime hours), the diurnal range is smaller in magnitude than temperature. It is also worth noting that the diurnal range is lower for Staten Island than for the Bronx or Queens, suggesting that degree of urbanization has a negative correlation with the diurnal range of $q$. Similar to surface temperature, the variability of $q$ is significantly lower during heat events ($ \sigma = \SI{2.14e-03}{\kilo\gram\per\kilo\gram} $) than during normal temperatures ($ \sigma = \SI{3.18e-03}{\kilo\gram\per\kilo\gram} $).
\\ \\
In the boundary layer, the positive $q$ anomalies subside in magnitude between 300 and \SI{600}{\meter}, but increase significantly in the mixed layer, especially during the late morning and early afternoon for all sites. As shown in Figure \ref{fig:extreme-heat-normal-comparison-contours-q}, the largest anomalies occur between 10:00 and 16:00 LST in the upper mixed layer. With regards to spatial variation in $q$, the highest anomalies occur over the Bronx and Queens, whereas Staten Island demonstrates a positive $q$ anomaly with lower magnitudes during extreme heat events. This correlates with the magnitude of the difference in surface quantities shown in Figure \ref{fig:extreme-heat-normal-0m-comparison-q}. 
\\ \\
With regards to time, mid-morning increases in anomaly values in the lower mixed layer (\texttildelow \SI{500}{\meter}) suggest enhanced vertical transport of moisture across all sites (see Figures \ref{fig:vertical_profiles-heat_wave-normal-q-bron}, \ref{fig:vertical_profiles-heat_wave-normal-q-quee}, \ref{fig:vertical_profiles-heat_wave-normal-q-stat}). Vertical profiles of $q$ across all locations show markedly higher $q$ values at the surface during extreme heat event with $\frac{dq}{dz}$ values increasing throughout the morning in the mixed layer while low-level $q$ values decrease, indicating vertical transport of moisture and drier low-level conditions during peak insolation.

@@

\subsubsection{Horizontal winds}
Extreme heat events coincided with an overall reduction of horizontal wind speeds ($U$) in the UBL. More specifically, the magnitude of $U$ during extreme heat events is similar in magnitude to $U$ during normal days with the exception of early morning hours and at upper levels of the UBL. As shown in Figure \ref{fig:extreme-heat-normal-comparison-contours-U}, modest reductions in $U$ ($-1.2 \leq \sigma \leq -0.4$) during extreme heat events are present throughout the UBL from early to mid-morning, with little differences throughout the rest of the day ($-0.4 \leq \sigma \leq 0.4$). Larger deviations between $U$ values are present at the top of the UBL where synoptic conditions become dominant.
\\ \\
Vertical profiles of $U$ for normal and extreme heat events provide a more detailed view of the differences in UBL structure. Across all sites, $U$ is similar throughout the UBL overnight, afternoon, and evenings. During early morning hours, extreme heat event $U$ values decrease by 25 to 50\% throughout the entire UBL (see Figures \ref{fig:vertical_profiles-heat_wave-normal-U-bron}, \ref{fig:vertical_profiles-heat_wave-normal-U-quee}, \ref{fig:vertical_profiles-heat_wave-normal-U-stat}), although both event types present a classical logarithmic wind profile, with surface friction effects present through \SI{500}{\meter}. Another phenomenon worth noting is the difference in $U$ profiles above \SI{2000}{\meter}; profiles of $U$ during extreme heat events are more consistent between sites and vertically than during normal days. This phenomenon may highlight the effect of synoptic meteorological conditions on $U$, as the UBL typically remains below \SI{2500}{\meter}. During extreme heat events, anticyclonic conditions produce more stable atmospheric conditions relative to normal days, resulting in less variability between heat events than during normal days.
\\ \\
Extreme heat events result in a southwesterly shift in $U$ throughout the UBL. This shift is present most evidently closer to the surface, as shown in Figures \ref{fig:wind_rose-bron}, \ref{fig:wind_rose-quee}, and \ref{fig:wind_rose-stat}, with winds at \SI{100}{\meter} coming primarily from the southwest quadrant. Figure \ref{fig:wind_rose-quee} shows that Queens also presents a secondary maximum with winds approaching from the south, which suggests effects from the Atlantic sea breeze (effects from the sea breeze will be further discussed in Section \ref{section:sea_breeze_effects}). At \SI{1000}{\meter}, the directionality of prevailing winds becomes more uniform between normal and extreme heat days, as winds primarily approach New York City from the west-southwest. The disparity in wind directions between 100 and \SI{1000}{\meter} suggests that localized wind fields play a major role in UBL dynamics at lower levels whereas synoptic-scale atmospheric conditions increasingly dominate with increasing height.
\\ \\
\subsubsection{Vertical motion}
On average, extreme heat events do not appear to produce significant changes in vertical velocity ($w$). Figure \ref{fig:extreme-heat-normal-comparison-100m-w} shows average diurnal profiles of $w$ at all locations at \SI{100}{\meter} above ground level, with similar mean values throughout the day between normal days and extreme heat events. During extreme heat events, however, the variability of $w$ is less in the early morning hours and greater in the evening. This phenomenon is also observed in vertical profiles of $w$ at all locations as shown in Figures \ref{fig:extreme-heat-normal-vertical_profiles-w-bron},\ref{fig:extreme-heat-normal-vertical_profiles-w-quee}, and \ref{fig:extreme-heat-normal-vertical_profiles-w-stat}. At all locations, overnight and morning profiles of $w$ (0:00 and 6:00 LST) show significantly lower variability in $w$ throughout the UBL with similar magnitudes of mean $w$. Despite similar means and deviations in the early afternoon (12:00 LST), evening profiles (18:00 LST) show significantly higher variability in $w$ below \SI{500}{\meter} for all sites, with the Bronx showing this occurrence extend through the UBL.

>>>>

\textbf{Observations and analysis of an urban boundary layer and sea breezes during extreme heat events}
Gabriel Rios \textsuperscript{1*, 2*}, Prathap Ramamurthy \textsuperscript{1, 2}

@@

Extreme heat presents a significant risk to human health and infrastructure in cities. Several studies have been conducted in the past several decades to understand the interaction between the synoptic-scale extreme heat events and local-scale urban heat island effects. However, observations of boundary layer characteristics during these periods have been relatively rare. Our current understanding of urban boundary layer structure is incomplete, particularly in coastal environments where the local climatology is highly influenced by land-sea thermal gradients. In this study, we analyze the evolution and structure of the urban boundary layer during regular and extreme heat periods with the goal of better understanding the effect of extreme heat and sea breezes on the boundary layer over a coastal urban area. Our analysis focuses on the New York City metropolitan area and relies on observations from vertical profilers (Doppler lidar, microwave radiometer) and quantities derived by analytical methods. Additionally, satellite and reanalysis data are used to supplement observational data. Extreme heat events present a mean peak surface temperature increase of \SI{7}{\kelvin}, an increase of site-averaged specific humidity at the surface by 39.4\%, and a marked southwesterly shift in winds at all sites. In addition, sea breeze events during heat extreme heat events are found to reduce temperatures and increase low-level moisture content from the early evening through nighttime hours, with strong variability between sites. The study also finds that extreme heat events unify horizontal wind directions throughout the boundary layer, and promote nocturnal onshore moisture transport.

@@

\begin{enumerate}
  \item How do UBL structure and dynamics depart from the climatology during extreme heat events?
  \item How do extreme heat events impact the transport of scalars?
  \item What effect does the sea breeze have on a coastal urban area during extreme heat events?
\end{enumerate}

@@

\subsection{Temperature}
On average, extreme heat events increase the temperature at the surface, as expected (see Figure \ref{fig:extreme_heat_normal_0m_comparison_T}). This is consistent across all observed locations in New York City, with the extreme heat event temperature exceeding normal temperatures by approximately 1-$\sigma$ over the entire day. An increase in the difference is observed during daytime hours, with the difference peaking in magnitude around 13:00 LST at the hottest time of day. The surface temperature variability is significantly lower during heat events (average $ \sigma = \SI{1.77}{\kelvin} $) than during normal temperatures (average $ \sigma = \SI{4.57}{\kelvin} $). There is little spatial variability between sites, with maximum average temperatures ranging from \SI{305.65}{\kelvin} in Queens to \SI{306.63}{\kelvin} in the Bronx. It is worth noting that there are areas in New York City that are located in more heavily urbanized areas than the observation sites (such as Midtown Manhattan and central Brooklyn), so it is likely that certain areas within the city have higher maximum temperatures. 
\\ \\
Above the surface, extreme heat events increase the temperature significantly over the lowest \SI{3000}{\meter} of the troposphere (see Figure \ref{fig:extreme-heat-normal-comparison-contours-theta}), with standardized anomalies of $\theta$ ranging from $\sigma = 0.99$ to $1.30$. The largest temperature anomalies shift from the surface layer in the mornings to span the entirety of the mixed layer. This is reflective of strong surface forcing resulting in convection through the mixed layer, as indicated by the formation of a late morning superadiabatic layer at all locations (Figures \ref{fig:vertical_profiles-heat_wave-normal-theta-bron}, \ref{fig:vertical_profiles-heat_wave-normal-theta-quee}, \ref{fig:vertical_profiles-heat_wave-normal-theta-stat}). 
\\ \\
These vertical profiles of $\theta$ suggest a degree of spatial variability in the UBL exists between locations. One instance of this spatial variability is vertical mixing; the Bronx site appears to have stronger vertical mixing as shown in Figure \ref{fig:vertical_profiles-heat_wave-normal-theta-bron}, as $\theta$ remains constant for a greater height than at the Queens and Staten Island locations, indicating a deeper mixed layer. This phenomenon is more pronounced during extreme heat events, as a distinct mixed layer is apparent in the Bronx during early (12:00 LST) and late (18:00 LST) afternoon hours. While a deepened mixed layer during extreme heat events is also visible for the other locations, the strength of vertical mixing in the Bronx is emphasized by persistent afternoon instability as shown by negative $\frac{d\theta}{dz}$ values between 500 and \SI{1000}{\meter} and a superadiabatic surface layer and 12:00 and 18:00 LST.

\subsection{Moisture}
On average, extreme heat events were found to increase the moisture at the surface, as indicated by the diurnal profiles of specific humidity ($q$) (see Figure \ref{fig:extreme-heat-normal-0m-comparison-q}). This is also consistent across all observed locations in New York City, with the extreme heat event temperature exceeding normal temperatures by approximately 1-$\sigma$ over the entire day. Although a distinct diurnal profiles exists ($q$ decreases during daytime hours), the diurnal range is smaller in magnitude than temperature. It is also worth noting that the diurnal range is lower for Staten Island than for the Bronx or Queens, suggesting that degree of urbanization has a negative correlation with the diurnal range of $q$. Similar to surface temperature, the variability of $q$ is significantly lower during heat events (average $ \sigma = \SI{2.14e-03}{\kilo\gram\per\kilo\gram} $) than during normal temperatures (average $ \sigma = \SI{3.18e-03}{\kilo\gram\per\kilo\gram} $). Queens shows exceptional variability in $q$, which may be attributed to the location of the observation site, which is adjacent to Flushing Meadows Corona Park (large open vegetated space), is surrounded by a medium-density urban area on all other sides, and is approximately \SI{4}{\kilo\meter} from Long Island Sound. 
\\ \\
In the boundary layer, the positive $q$ anomalies subside in magnitude between 300 and \SI{600}{\meter}, but increase significantly in the mixed layer, especially during the late morning and early afternoon for all sites. As shown in Figure \ref{fig:extreme-heat-normal-comparison-contours-q}, the largest anomalies occur between 10:00 and 16:00 LST throughout the mixed layer. With regards to spatial variation in $q$, Staten Island demonstrates a strong positive anomaly overnight through the early morning near the surface, indicating elevated low-level moisture transport during extreme heat events, whereas the Bronx and Queens demonstrate a similar phenomenon with a lesser anomaly magnitude. All sites show significant positive $q$ anomalies throughout the day, with the strongest anomaly signal starting in the low-levels throughout the morning and transitioning to the mixed later by mid-afternoon. This trend suggests that the increase in nocturnal low-level moisture corresponds to increased UBL moisture content due to strong vertical mixing throughout the daytime.
\\ \\
This is supported by Figures \ref{fig:vertical_profiles-heat_wave-normal-q-bron}, \ref{fig:vertical_profiles-heat_wave-normal-q-quee}, \ref{fig:vertical_profiles-heat_wave-normal-q-stat}, where vertical profiles of $q$ across all locations show markedly higher $q$ values at the surface during extreme heat events with $\frac{dq}{dz}$ values increasing throughout the morning in the mixed layer while low-level $q$ values decrease, indicating vertical transport of moisture and drier low-level conditions during peak insolation. The strong vertical mixing of $q$ can be observed at all sites, where late morning and early afternoon $\frac{dq}{dz}$ values are greater during extreme heat events than normal days. An example can be seen in the Bronx, where $\frac{dq}{dz} > 0$, indicating very efficient vertical moisture transport.

@@

\subsubsection{Horizontal winds}
Extreme heat events coincided with an overall reduction of horizontal wind speeds ($U$) in the UBL. More specifically, the magnitude of $U$ during extreme heat events is similar in magnitude to $U$ during normal days with the exception of early morning hours and at upper levels of the UBL. As shown in Figure \ref{fig:extreme-heat-normal-comparison-contours-U}, modest reductions in $U$ ($-1.2 \leq \sigma \leq -0.4$) during extreme heat events are present throughout the UBL from early to mid-morning, with little differences throughout the rest of the day ($-0.4 \leq \sigma \leq 0.4$). Larger deviations between $U$ values are present at the top of the UBL where synoptic conditions become dominant.
\\ \\
Vertical profiles of $U$ for normal and extreme heat events provide a more detailed view of the differences in UBL structure. Across all sites, $U$ is similar throughout the UBL overnight, afternoon, and evenings. During early morning hours, extreme heat event $U$ values decrease by 25 to 50\% throughout the entire UBL (see Figures \ref{fig:vertical_profiles-heat_wave-normal-U-bron}, \ref{fig:vertical_profiles-heat_wave-normal-U-quee}, \ref{fig:vertical_profiles-heat_wave-normal-U-stat}), although both event types present a classical logarithmic wind profile, with surface friction effects present through \SI{500}{\meter}. Another phenomenon worth noting is the difference in $U$ profiles above \SI{2000}{\meter}; profiles of $U$ during extreme heat events are more consistent between sites and vertically than during normal days. This phenomenon may highlight the effect of synoptic meteorological conditions on $U$, as the UBL typically remains below \SI{2500}{\meter}. During extreme heat events, anticyclonic conditions produce more stable atmospheric conditions relative to normal days, resulting in less variability between heat events than during normal days.
\\ \\
Extreme heat events result in a southwesterly shift in $U$ throughout the UBL. This shift is present most evidently closer to the surface, as shown in Figures \ref{fig:wind_rose-bron}, \ref{fig:wind_rose-quee}, and \ref{fig:wind_rose-stat}, with winds at \SI{100}{\meter} coming primarily from the southwest quadrant. Figure \ref{fig:wind_rose-quee} shows that Queens also presents a secondary maximum with winds approaching from the south, which suggests effects from the Atlantic sea breeze (effects from the sea breeze will be further discussed in Section \ref{section:sea_breeze_effects}). At \SI{1000}{\meter}, the directionality of prevailing winds becomes more uniform between normal and extreme heat days, as winds primarily approach New York City from the west-southwest. The disparity in wind directions between 100 and \SI{1000}{\meter} suggests that localized wind fields play a major role in UBL dynamics at lower levels whereas synoptic-scale atmospheric conditions increasingly dominate with increasing height.

\subsubsection{Vertical motion}
On average, extreme heat events do not appear to produce significant changes in vertical velocity ($w$). Figure \ref{fig:extreme-heat-normal-comparison-100m-w} shows average diurnal profiles of $w$ at all locations at \SI{100}{\meter} above ground level, with similar mean values throughout the day between normal days and extreme heat events. During extreme heat events, however, the variability of $w$ is less in the early morning hours and greater in the evening. This phenomenon is also observed in vertical profiles of $w$ at all locations as shown in Figures \ref{fig:extreme-heat-normal-vertical_profiles-w-bron},\ref{fig:extreme-heat-normal-vertical_profiles-w-quee}, and \ref{fig:extreme-heat-normal-vertical_profiles-w-stat}. At all locations, overnight and morning profiles of $w$ (0:00 and 6:00 LST) show significantly lower variability in $w$ throughout the UBL with similar magnitudes of mean $w$. Despite similar means and deviations in the early afternoon (12:00 LST), evening profiles (18:00 LST) show significantly higher variability in $w$ below \SI{500}{\meter} for all sites, with the Bronx showing this occurrence extend through the UBL.
